Processing large surveys with package MRP-KIT

### 2021-06-30 case-study/Munging_CCES.Rmd
checkpointing case study

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, message=FALSE, error=FALSE, warning=FALSE, comment=NA, out.width='95%', collapse=TRUE)
```

<style type="text/css">
.table { width: 40%; }
div.quote-container blockquote {  font-size: small; }
</style>

The goal of this case study is to show how the MRP-KIT package
can facilitate analysis of a large-scale survey sample.
By large-scale, we mean surveys which have both a rich set
of demographics and a large number of responses.
The richer the set of demographics, the larger the number
of possible sub-populations of interest; no matter how large
or representative the survey, the combinatorics of the demographic
categories are such that it is unlikely that the sample
is representative for all subpopulations of interest.
In order to correct this imbalance,
we post-stratify to a truly representative
sample drawn from the general population.
This, in turn, requires obtaining a general population sample;
one which is at least as large and which contains all the
demographics of interest.

In this case study, the survey sample is one year's worth of data
from is Cooperative Congressional Election Study (CCES),
a bi-annual survey of of 60,000 American adults from
all 50 states and the District of Columbia.
The CCES is designed to capture 
the views and experiences of Amercians with respect to Congress
and congressional (and presidential) elections.
While this survey sample is representative at the national level
it is not large enough to provide a representative sample for sub-populations
of interest.
The population sample is the 2018 American Community Survey, the
annual micro-census conducted by the US Census Bureau.  It is designed
to be a representative sample of the US population as a whole; data is
collected from 3 million households.

@@

The 2018 CCES Common Content Dataset is 
available from the Harvard Dataverse project:
https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi%3A10.7910/DVN/ZSBZ7K
The raw survey data is distributed in several from the Harvard Dataverse
as  [cces18_common_vv](https://dataverse.harvard.edu/file.xhtml?persistentId=doi:10.7910/DVN/ZSBZ7K/C1YVSF&version=6.0).
Although the tabular versions of the data file are smaller, we download the
Stata file, as it contains metadata on all the questions; this is critical to
being able to align the CCES sample to the US Census population sample.
We also download the [2018 CCES Guide](https://dataverse.harvard.edu/file.xhtml?persistentId=doi:10.7910/DVN/ZSBZ7K/WZWCZ1&version=6.0)
which describes the survey questions, labels, and responses.

We load the data into R using the [`readstata13`](https://sjewo.github.io/readstata13/) library function
[`read.dta13`](https://sjewo.github.io/readstata13/reference/read.dta13.html) which returns a data.frame with attributes.
```{r}
library(readstata13)
cces_2018 = read.dta13("cces18_common_vv.dta", convert.factors = T, nonint.factors = T, generate.factors=T)
```

@@

The Common Content portion of the survey consists of questions asked of all subjects.
Profile questions provide demographic information,
and opinion questions provide outcomes of interest.
The demographic variables include
gender, education level, employment status, race, and geographic location.
```{r}
table(cces_2018$gender[,drop=TRUE])
table(cces_2018$educ[,drop=TRUE])
table(cces_2018$race[,drop=TRUE])
```
The political opinion questions on the survey are labelled "Abortion", "Immigration", "Taxes", "Health Care".
```{r}
attr(cces_2018, "var.labels")[which(startsWith(attr(cces_2018, "var.labels"), "Health Care"))[1]]
health_care_1 <- which(startsWith(attr(cces_2018, "var.labels"), "Health Care"))[1]
table(cces_2018[[health_care_1]][, drop=TRUE])
```

Of the 525 columns in the survey, we are interested in the set of political opinion questons
and a subset of the demographics.  Although not strictly necessary, we create a new dataframe
which contains just these columns of interest.
The survey metadata includes column names, and question and response labels.
We use the column names to select demographics and column labels to select political opinion questions.
```{r}
col_labels <- attr(cces_2018, "var.labels")
col_names <- attr(cces_2018, "names")
demo_cols <- c(1, which(col_names == "birthyr"),
              which(col_names == "gender" ),
              which(col_names == "educ"),
              which(col_names == "race"),
              which(col_names == "hispanic"),
              which(col_names == "marstat"),
              which(col_names == "region"),
              which(col_names == "employ"),
              which(col_names == "hadjob"),
              which(col_names == "inputstate"))
outcome_cols <- c(which(startsWith(col_labels, "Abortion --")),
                 which(startsWith(col_labels, "Immigration --")),
                 which(startsWith(col_labels, "Taxes --")),
                 which(startsWith(col_labels, "Health Care --")),
                 which(startsWith(col_labels, "President 20")))
col_labels[demo_cols]
col_labels[outcome_cols]
keep_cols <- sort(c(demo_cols, outcome_cols))
cces_2018_subset <- subset(cces_2018[keep_cols])
```

@@

The CCES data is roughly representative of the state-level political sentiment as a whole; but if we seek to understand certain subsets of the population,
the sample may well be unbalanced.
There are only 60,000 survey responses,
but there are over 20 million possible demographic subsets of interest,
given all possible combinations of gender, education,
race, marital status, employment, birthyear, and state of residence (51).

```{r}
cat(format((length(table(cces_2018_subset$gender[,drop=TRUE])) *
length(table(cces_2018_subset$educ[,drop=TRUE])) *
length(table(cces_2018_subset$race[,drop=TRUE])) *
length(table(cces_2018_subset$marstat[,drop=TRUE])) *
length(table(cces_2018_subset$employ[,drop=TRUE])) *
length(table(cces_2018_subset$birthyr[,drop=TRUE])) *
length(table(cces_2018_subset$inputstate[,drop=TRUE]))), big.mark=",",scientific=FALSE))
```

In order to correct this imbalance, we must poststratify to the general population.

@@

The US Census data is the largest and most representative population sample we have acces to.
A full census is conducted every 10 years.
The American Community Survey (ACS) is a micro-census conducted every year by the Census Bureau
over 1% of the households in the US, i.e., it has 3 million respondants (as opposed to 300 million in the full census).

@@

* As even the US Census microdata files are extremely large, the IPUMS service queues this request for offline processing.
When the data has been assembled, IPUMS sends an email with a link to the download page, shown below.

@@

You must download the data file and the DDI file for the codebook, as well as the R command file.
The DDI file is an XML file which documents your data extract.  If you click on this link,
the DDI file displays in the browser; you must download this file as XML.
The IPUMS website provides [detailed instructions](https://usa.ipums.org/usa/extract_instructions.shtml).
Both the data file and xml download file have the same base name;
if you rename them, you must rename both of them accordingly.

@@

The columns of this tibble are  [labelled](https://haven.tidyverse.org/reference/labelled.html) vectors.
Labelled vectors can have text labels associated with specific values.
To see how this works, we compare the values and labels for `AGE`.
The 2018 ACS data spans ages 0 through 96,
of these, only ages 0 and 90 have labels.
```{r}
length(table(acs_2018_subset$AGE))
max(acs_2018_subset$AGE)
```

The function [`as_factor`](https://haven.tidyverse.org/reference/as_factor.html)
allows us to treat labelled vectors as factors.

@@

We need to align our demographics of interest; we will do so column-by-column.

We note that information which must be aggregated from combinations of answers
given to a series of questions in order to match demographics in the ACS
is beyond the scope of what can be done with MRP-KIT, e.g.,
it doesn't provide functions for combining survey variables.

Also problematic: creating aggregate categories, e.g., binning age by decade.

#### Variable CCES "gender" / ACS "sex"

Both the CCES and ACS samples have variables for the sex of the respondant:
in the CCES this is named "gender" and in the ACS this is named "SEX".
In both, the response labels are "Male", "Female".
Creating a mapping between the two is straightforward:
we create a new SurveyQuestion object with name "sex",
then create a mappng from the CCES to the ACS,
both for column names and response labels
where for both mappings, the first element
is the survey sample name/label, and the second element
is the population sample name/label.

```{r}
library(mrpkit)
names(table(cces_2018_subset$gender))
print(ipums_val_labels(acs_2018_subset$SEX))
q_sex <- SurveyQuestion$new(name="sex",
      col_names=c("gender", "SEX"),
      values_map=list("Male" = "Male","Female" = "Female"))
print(q_sex)
```

#### Variable CCES "inputstate" / ACS "STATEFIP"

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, message=FALSE, error=FALSE, warning=FALSE, comment=NA, out.width='95%', collapse=TRUE, prompt=TRUE)
```

<style type="text/css">
.table { width: 85%; }
div.quote-container blockquote {  font-size: small; }
</style>

Multi-level Regresssion and Postrativication (MRP)
is a two-step process for correcting model estimates
obtained from a survey sample when the survey population is not
the same as the target population over which we wish to do inference.
The first step is the doing a multilevel regression on the survey sample;
the second step is poststratification to the target population.
Conceptually, MRP is relatively simple.
In practice, it requires more than twice as much data processing.
In addition to managing the survey data, you must also obtain
and manage the target population data, in addition you must
develop a mapping between information in these two datasets.
In order to produce verifiable, replicable results, it is
critical that this mapping be documented and scripted.

The goal of this case study is to show how the MRP-KIT package
can facilitate this process for large-scale surveys.
By large-scale surveys, we mean surveys which have both
a large number of respondants and detailed demographic profiles for each respondant.
The larger the number of demographics, the larger the number
of possible sub-populations of interest and the more difficult it becomes
to collect a representative sample for all such subpopulations,
thus even well-designed surveys can require use of MRP.

In this case study, the survey sample is one year's worth of data
from is Cooperative Congressional Election Study (CCES),
a bi-annual survey of of 60,000 American adults from
all 50 states and the District of Columbia.
The CCES is designed to capture 
the views and experiences of Amercians with respect to Congress
and congressional (and presidential) elections.
While this survey sample is representative at the national level
it is not large enough to provide a representative sample for sub-populations
of interest.
The population sample is the 2018 American Community Survey (ACS), the
annual micro-census conducted by the US Census Bureau.  It is designed
to be a representative sample of the US population as a whole; data is
collected from 3 million households.

@@

The 2018 CCES Common Content Dataset is 
available from the Harvard Dataverse project:
https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi%3A10.7910/DVN/ZSBZ7K
The raw survey data is distributed in several formats from the Harvard Dataverse
as  [cces18_common_vv](https://dataverse.harvard.edu/file.xhtml?persistentId=doi:10.7910/DVN/ZSBZ7K/C1YVSF&version=6.0).
While the tabular versions of the data file are smaller, the Stata version `cces18_common_vv.dta`
contains metadata on all the questions: the question name, label, description, response labels and values.
The corresponding 
[2018 CCES Guide](https://dataverse.harvard.edu/file.xhtml?persistentId=doi:10.7910/DVN/ZSBZ7K/WZWCZ1&version=6.0)
provides a human-readable version of this metadata as well.

```{r ccesDownload, echo=FALSE, out.width = "75%", fig.cap="Harvard Dataverse download page"}
library(knitr)
include_graphics("dataverse_cces_2018_download.png")
```

We download the Stata file because we will need the metatdata to map the CCES sample to the ACS using MRP-KIT,
then use the [`readstata13`](https://sjewo.github.io/readstata13/) library function
[`read.dta13`](https://sjewo.github.io/readstata13/reference/read.dta13.html) to load the Stata file into R.
The result is an R data.frame with attributes containing the metadata.
```{r}
library(readstata13)
cces_2018 = read.dta13("cces18_common_vv.dta", convert.factors = T, nonint.factors = T, generate.factors=T)
```

@@

The Common Content portion of the survey consists of questions asked of all subjects.
Profile questions provide demographic information,
and opinion questions provide outcomes of interest.
The demographic variables include
gender, education level, race, and geographic location, among others.
```{r}
table(cces_2018$gender[,drop=TRUE])
table(cces_2018$educ[,drop=TRUE])
table(cces_2018$race[,drop=TRUE])
```
The political opinion questions on the survey are labelled "Abortion", "Immigration", "Taxes", "Health Care".
```{r}
attr(cces_2018, "var.labels")[which(startsWith(attr(cces_2018, "var.labels"), "Health Care"))[1]]
health_care_1 <- which(startsWith(attr(cces_2018, "var.labels"), "Health Care"))[1]
table(cces_2018[[health_care_1]][, drop=TRUE])
```

Of the 525 columns in the survey, we are interested in the set of political opinion questons
and a subset of the demographics.  Although not strictly necessary, we create a new dataframe
which contains just these columns of interest.
The survey metadata includes column names, and question and response labels.
We use the column names to select demographics and column labels to select political opinion questions.
```{r}
col_labels <- attr(cces_2018, "var.labels")
col_names <- attr(cces_2018, "names")
demo_cols <- c(1, which(col_names == "birthyr"),
              which(col_names == "gender" ),
              which(col_names == "educ"),
              which(col_names == "race"),
              which(col_names == "hispanic"),
              which(col_names == "marstat"),
              which(col_names == "inputstate"))
outcome_cols <- c(which(startsWith(col_labels, "Abortion --")),
                 which(startsWith(col_labels, "Immigration --")),
                 which(startsWith(col_labels, "Taxes --")),
                 which(startsWith(col_labels, "Health Care --")),
                 which(startsWith(col_labels, "President 20")))
col_labels[demo_cols]
col_labels[outcome_cols]
keep_cols <- sort(c(demo_cols, outcome_cols))
cces_2018_subset <- subset(cces_2018[keep_cols])
```

@@

The CCES data is roughly representative of the state-level political sentiment as a whole,
but if we seek to understand certain subsets of the population,
the sample may well be unbalanced.
There are only 60,000 survey responses,
but there are over 2 million
possible combinations of gender, education,
race, marital status, birthyear, and state of residence.
If we bin ages by decade, there are a mere 200,000 combinations,
binning states by region further reduces this to 16,000, etc., etc.
However, if we wish to consider state-level subpopulations of interest,
the survey data is likely to be sparse and not representative of the general population.
To address the former we do multi-level modeling; for the latter, poststratification.

@@

The US Census data is the largest representative population sample that we have access to.
A full census is conducted every 10 years.
The American Community Survey (ACS) is a micro-census conducted every year by the Census Bureau
over 1% of the households in the US, i.e., it has 3 million respondants (as opposed to 300 million in the full census).

@@

The IPUMS service queues this request for offline processing.
When the data has been assembled, IPUMS sends an email with a link to the download page, shown below.

@@

You must download the data file and the DDI file for the codebook, as well as the R command file.
The IPUMS website provides [detailed instructions](https://usa.ipums.org/usa/extract_instructions.shtml) on how to do this.
The DDI file is an XML file which documents your data extract.  If you click on this link,
the DDI file displays in the browser; you must download this file as XML.
Both the data file and xml download file have the same base name;
if you rename them, you must rename both of them accordingly.

@@

The columns of this tibble are  [labelled](https://haven.tidyverse.org/reference/labelled.html) vectors.
Labelled vectors can have text labels associated with specific values.
To see how this works, we compare the values and labels for `AGE`.
The 2018 ACS data spans ages 0 through 96,
of these, only ages 0 and 90 have labels.
The function [`ipums_val_labels`](https://tech.popdata.org/ipumsr/reference/ipums_var_info.html)
allows us to inspect these labels.
```{r}
length(table(acs_2018_subset$AGE))
ipums_val_labels(acs_2018_subset$AGE)
```

The function [`as_factor`](https://haven.tidyverse.org/reference/as_factor.html)
allows us to treat labelled vectors as factors.
See the vignette https://tech.popdata.org/ipumsr/reference/ipums_var_info.html for more details.

@@

We use MRP-KIT to map demographics of interest on a column-by-column basis.^[
If information is spread across multiple columns in either the survey sample
or the target population sample, it must be combined into a single column's
worth of information as MRP-KIT doesn't provide functions to do this.
*Note: all pre-processing should be documented, scripted, and **tested**!*]

In the following sections we examine each variable in turn.
This is a tedious process, therefore, we provide the following
summary of the demographic construct and the kinds of challenge each mapping presents

Construct | CCES | ACS | Challenge
--------- | ---- | --- | ---------
Sex | gender | SEX | (none)
State | inputstate | STATEFIP | specify 1-1 mapping programatically
Marital status | marstat | MARST | choice of mapping
Education | educ | EDUC | choice of mapping
Race/Hispanic ethnicity | race, hispan | RACE, HISPAN | cannot align CCES race to ACS RACE
Age | birthyr | AGE | different units, too fine-grained

#### Trivial mapping: respondant sex, 2 categories

The CCES and ACS samples record the sex of the respondant;
in the CCES this is named "gender" and in the ACS this is named "SEX".
In both, the response labels are "Male", "Female".
Creating a mapping between the two is straightforward:
we create a new SurveyQuestion object with name "sex",
then create a mappng from the CCES to the ACS,
both for column names and response labels
where for both mappings, the first element
is the survey sample name/label, and the second element
is the population sample name/label.

```{r}
library(mrpkit)
names(table(cces_2018_subset$gender))
ipums_val_labels(acs_2018_subset$SEX)
q_sex <- SurveyQuestion$new(name="sex",
      col_names=c("gender", "SEX"),
      values_map=list("Male" = "Male","Female" = "Female"))
q_sex
```

#### 1-1 mapping on a long list: state of residence, 51 categories

@@

#### Choice of mapping: education, marital status

The CCES and ACS samples record the education level of the respondant
in variables `educ` and `EDUC`, respectively.
Both the CCES and ACS record the highest level of education completed,
but the CCES records the degree obtained whereas the ACS records the year completed.
```{r}
names(table(cces_2018_subset$educ))
ipums_val_labels(acs_2018_subset$EDUC)
```
We choose to align these categories as follows:

CCES | ACS
---- | ---
"No HS" | "Nursery school to grade 4"
"No HS" | "Grade 5, 6, 7, or 8"
"No HS" | "Grade 9"
"No HS" | "Grade 10"
"No HS" | "Grade 11"
"High School Graduate" | "Grade 12"
"Some College" | "1 year of college"
"2-year" | "2 years of college"
"2-year" | "3 years of college"
"4-year" | "4 years of college"
"Post-grad" | "5+ years of college"
"skipped" | "N/A or no schooling"
"not asked" | "N/A or no schooling"

Alternative mappings are equally viable.
With MRP-KIT, the choice of mappings is recorded and
easily accessible.
```{r}
educ_mapping = list(
    "No HS" = "Nursery school to grade 4",
    "No HS" = "Grade 5, 6, 7, or 8",
    "No HS" = "Grade 9",
    "No HS" = "Grade 10",
    "No HS" = "Grade 11",
    "High School Graduate" = "Grade 12",
    "Some College" = "1 year of college",
    "2-year" = "2 years of college",
    "2-year" = "3 years of college",
    "4-year" = "4 years of college",
    "Post-grad" = "5+ years of college",
    "skipped" = "N/A or no schooling",
    "not asked" = "N/A or no schooling"
)
q_educ <- SurveyQuestion$new(name = "educ",
    col_names = c("educ", "EDUC"),
    values_map= educ_mapping)
q_educ$name()
q_educ$col_names()
head(q_educ$values())
```

As with eduction, the CCES and ACS samples record the marital status
of the respondant differently and we have to use our judgement as to how best
to align the two.
```{r}
names(table(cces_2018_subset$marstat))
ipums_val_labels(acs_2018_subset$MARST)
```

CCES | ACS
---- | ---
"Married" | "Married, spouse present"
"Married" | "Married, spouse absent"
"Separated" | "Separated"
"Divorced" | "Divorced"
"Widowed" | "Widowed"
"Never married" | "Never married"
"Domestic / civil partnership" | "Never married"
"skipped" | "Never married"
"not asked" | "Never married"

We note that for the 2018 CCES survey, the marital status of all respondants is known; nonetheless,
we provide a complete mapping.

```{r}
marstat_mapping = list(
    "Married" = "Married, spouse present",
    "Married" = "Married, spouse absent",
    "Separated" = "Separated",
    "Divorced" = "Divorced",
    "Widowed" = "Widowed",
    "Never married" = "Never married",
    "Domestic / civil partnership" = "Never married",
    "skipped" = "Never married",
    "not asked" = "Never married"
)
q_marstat <- SurveyQuestion$new(name = "marstat",
    col_names = c("marstat", "MARST"),
    values_map= marstat_mapping)
q_marstat$name()
q_marstat$col_names()
head(q_marstat$values())
```

#### Information split across survey columns:  race and ethnicity.

We ignore the politics of race and racial categories
and instead focus on how best to map the CCES information to the ACS information.
Both surveys have multiple variables which provide information about race and ethnicity,
and both record race and hispanic origin separately.
However the CCES treats hispanic origin as race as well as ethnic origin; the ACS does not.
```{r}
names(table(cces_2018_subset$race))
names(table(cces_2018_subset$hispan))
ipums_val_labels(acs_2018_subset$RACE)
ipums_val_labels(acs_2018_subset$HISPAN)
```
In this case, munging the raw survey data is unavoidable.
For each survey we create a new variable which blends race and hispanic ethnicity.
We are scripting everyting, so we add this to our processing scripts.
